### 2014-10-10 paper/include/background.tex
corrected

The memory controller (MC) normally issues refresh operations periodically at an interval of $64\:ms$ for each row, an interval that has been constant for the latest DRAM specifications [source]. This interval is conservatively halved if the DRAM exceeds the specified working temperature of $85^{\circ}C$.

@@

The implementation of a DRAM refresh cycle can be made in two different ways, to either affect one row or many. The first is called RAS-only refresh (ROR) where refresh cycle refreshes one targeted row. The address for the targeted row has to be put on the bus and it is the MC that makes sure that a row are refreshed within the $64\:ms$ interval. The second is called CAS before RAS refresh (CBR). In this scheme the memory module has an internal address counter that displays which row that shall be refreshed. On a refresh cycle, the row stated by the internal address counter is refreshed and the counter is then incremented. The advantage with this method is that the address is never put on the bus, which gives a lower power consumption and less bus usage than ROR.

When using CBR, all banks in a rank are blocked for normal accesses. All banks are also blocked when ROR is used due to DRAM chip design, however some DRAM chips support per-bank refresh commands that makes it possible to only block the specific bank which is getting refreshed.

@@

The trends of DRAM chips is that they become larger and larger. In practice, this is realized to a high degree by adding additional rows in the banks. The rows could instead - or also - been made longer by adding DRAM cells, but this solution is not favorable as it increases the power consumption per row access.

As refresh operations are made by row, the refresh operations will increase linear with the device capacity. In turn, the throughput loss increases exponentially as in \reffig{fig:dram_throughput}. Additional refresh operations gives larger power consumption and the refreshes becomes a growing portion of the total DRAM energy consumption as seen in \reffig{fig:dram_power}. When the device capacity reaches 64 GB the refresh power nearly constitute half of the total consumption, an increase from 15\% if the device capacity is 8 GB.

The conclusion drawn from \reffig{fig:dram_data_proj} is that it will be increasingly important to mitigate the throughput loss and DRAM power consumption of future systems. One way to decrease these negative trends of DRAM scaling is to optimize the DRAM refresh. However, a reduction of the refreshes made has no linear relationship to the reduction of refresh energy consumption. The energy reduction depends on the state of the bank in which the targeted row resides; if the targeted row is closed more energy will be consumed than if the targeted row would have been open.

>>>>

The memory controller (MC) normally issues refresh operations periodically at an interval of $64\:ms$ for each row, an interval that has been constant for the latest DRAM specifications [source] and called \textit{maximum refresh period}. This interval is conservatively halved if the DRAM exceeds the specified working temperature of $85^{\circ}C$.

@@

The implementation of a DRAM refresh cycle can be made in two different ways, to either affect one row or many. The first is called \textit{RAS-only refresh} (ROR) where a refresh cycle refreshes one targeted row. To target one row, the row's address is put on the bus, and the MC have the responsibility that a row is refreshed within the maximum refresh period. The second is called \textit{CAS before RAS refresh} (CBR). In this scheme, the memory module have an internal address counter that displays which row that will be refreshed. On a refresh cycle, the row stated by the internal address counter is refreshed and the counter is then incremented. Contrary to ROR, CBR do not need provided addresses on the bus, which results in a lower power consumption and less bus usage. % During CBR, a row in each bank is updated, right? if so. write that in the text.

When using CBR, all banks in a rank are blocked for normal accesses. All banks are also blocked when ROR is used due to DRAM chip design, however some DRAM chips support per-bank refresh commands that makes it possible to only block the specific bank being refreshed.

@@

The trends in DRAM chips is to increase capacity with each generation. In practice, this is realized to a high degree by adding additional rows in the banks. The rows could instead - or also - been made longer by adding DRAM cells, but this solution is not favorable as it increases the power consumption per row access.

As refresh operations are made by row, the refresh operations will increase linearly with the device capacity. In turn, the throughput loss increases exponentially, as shown in \reffig{fig:dram_throughput}. Additional refresh operations give larger power consumption and the refreshes become a growing portion of the total DRAM energy consumption, as shown in \reffig{fig:dram_power}. When the device capacity reaches $64\:GB$, the refresh power nearly constitutes half of the total consumption, an increase from 15\% if the device capacity is 8 GB.

The conclusion to draw from \reffig{fig:dram_data_proj} is that it will be increasingly important to mitigate the throughput loss and DRAM power consumption of future systems. One way to decrease these negative trends of DRAM scaling is to optimize the DRAM refresh. However, a reduction in the number of refreshes made has no linear relationship to the reduction in refresh energy consumption. The energy reduction depends on the state of the bank in which the targeted row resides; if the targeted row is closed more energy will be consumed than if the targeted row would have been open.
